---
title: "polyAprobe analysis and figure generation"
author: "darach"
date: "`r Sys.Date()`"
---

```{r}
library(flowCore)
library(tidyverse)
```


```{r,cache=T}

bdf <- read.csv("polyeh/bdf160608.csv")
udf <- read.csv("polyeh/exp148160608.csv")

head(bdf)

head(udf)
```


```{r,cache=T}
g<-ggplot(subset(bdf,fix==120&probe==T))+theme_bw()+
  aes(x=media,y=log10(FL1.A))+
  ylab("log10( polyA signal, per cell )")+xlab("")+
  scale_x_discrete(labels=c("N-limited (Proline)","Rich media (YPD)"))+
  geom_boxplot(outlier.size=0.1,width=0.3)+#+coord_cartesian(ylim=c(0.1,7.0))
  theme(axis.text.x=element_text(angle=0,color="black"))
g
```
```{r,cache=F}
ggsave("../img/polya_ypdnlim.png",g,width=4,height=4)
```
```{r,cache=T}
g<-ggplot(subset(bdf,fix==120))+theme_bw()+
  aes(x=factor(media):factor(probe),y=(FL1.A))+#/FSC.A))+
  ylab("polyA signal, per cell")+xlab("")+
  scale_x_discrete(labels=c("RNAsed Proline-grown","N-limited (Proline)",
    "RNAsed YPD-grown","Rich media (YPD)"))+
  geom_boxplot(outlier.size=0.1,width=0.3)+coord_cartesian(ylim=c(0.0,1e5))+
  theme(axis.text.x=element_text(angle=90))
g
```

```{r,cache=F}
ggsave("../img/polya_ypdnlim_controls.png",g,width=4,height=4)
```

```{r,cache=T}
g2<-ggplot(subset(bdf,fix==120&probe==T))+theme_bw()+
  aes(x=factor(media):factor(probe),y=(FL1.A))+#/FSC.A))+
  ylab("polyA signal, per cell")+xlab("")+
  scale_x_discrete(labels=c("0.15 rate, \nN-limited (Proline)",
    "~0.45 rate,\nRich media (YPD)"))+
  geom_violin(adjust=0.1)+
  geom_boxplot(outlier.size=0.1,width=0.1)+
  scale_y_log10(limits=c(1e3,3e5))+
#  coord_cartesian(ylim=c(0.0,1.5e5))+
#  theme(axis.text.x=element_text(angle=90))
  theme(axis.text.x=element_text(angle=0,color="black"))
g2
```
```{r,cache=F}
ggsave("../img/polya_ypdnlim_prezi.png",g2,width=4,height=4)
```
```{r,cache=T}

g<-ggplot(subset(bdf,fix==120))+theme_bw()+
  aes(x=factor(media):factor(probe),y=log10(FL1.A))+
  ylab("log10( polyA signal, per cell )")+xlab("")+
  scale_x_discrete(labels=c("RNAsed Proline-grown","N-limited (Proline)",
    "RNAsed YPD-grown","Rich media (YPD)"))+
  geom_boxplot(outlier.size=0.1,width=0.3)+coord_cartesian(ylim=c(1.0,6.0))
g

subset(bdf,fix==120&probe==T)%>%group_by(media)%>%summarize(fl1a=list(FL1.A))%>%{ks.test(unlist(.[1,2][[1]]),unlist(.[2,2][[1]]))}

subset(bdf,fix==120&probe==T)%>%group_by(media)%>%summarize(fl1a=list(FL1.A))%>%{wilcox.test(unlist(.[1,2][[1]]),unlist(.[2,2][[1]]))}

subset(bdf,fix==120&probe==T)%>%group_by(media)%>%summarize(fl1a=mean(FL1.A))%>%{.[2,2]/.[1,2]}
subset(bdf,fix==120&probe==T)%>%group_by(media)%>%summarize(fl1a=median(FL1.A))%>%{.[2,2]/.[1,2]}


u<-ggplot(subset(udf,probe==T))+theme_bw()+
  aes(x=as.numeric(sample),group=sample,y=log10(FL1.A))+
  ylab("log10( polyA signal, per cell )")+
  xlab("Minutes after upshift")+geom_boxplot(outlier.size=0.1,width=10)+
  coord_cartesian(ylim=c(3.7,5.25),xlim=c(-10,250))+
  annotate(geom="text",x=-13,y=4.60,angle=90,label="Nlim Proline, +400uM Q")
u
```
```{r,cache=F}
ggsave("../img/polya_upshift.png",u,width=5,height=3)
ggsave("../img/polya_upshift_prezi.png",u+
  coord_cartesian(ylim=c(3.7,5.25),xlim=c(-10,130))
,width=4,height=4)
```
```{r,cache=T}


subset(udf,sample%in%c(0,123))%>%group_by(sample)%>%summarize(fl1a=mean(FL1.A))%>%{.[2,2]/.[1,2]}
subset(udf,sample%in%c(0,123))%>%group_by(sample)%>%summarize(fl1a=median(FL1.A))%>%{.[2,2]/.[1,2]}



g<-ggplot(subset(bdf,fix==120&probe==T))+theme_bw()+
  aes(x=media,y=(FL1.A/FSC.A))+
  ylab("polyA signal / FSC , per cell")+xlab("")+
  scale_x_discrete(labels=c("N-limited (Proline)","Rich media (YPD)"))+
  geom_boxplot(outlier.size=0.1,width=0.3)+coord_cartesian(ylim=c(0,.75))
g

g<-ggplot(subset(bdf,fix==120))+theme_bw()+
  aes(x=factor(media):factor(probe),y=(FL1.A/FSC.A))+
  ylab("polyA signal / FSC, per cell")+xlab("")+
  scale_x_discrete(labels=c("RNAsed Proline-grown","N-limited (Proline)",
    "RNAsed YPD-grown","Rich media (YPD)"))+
  geom_boxplot(outlier.size=0.1,width=0.3)+coord_cartesian(ylim=c(0.0,.75))
g

u<-ggplot(subset(udf,probe==T))+theme_bw()+
  aes(x=as.numeric(sample),group=sample,y=(FL1.A/FSC.A))+
  ylab("polyA signal / FSC, per cell")+xlab("")+
  xlab("Minutes after upshift")+geom_boxplot(outlier.size=0.1,width=10)+
  coord_cartesian(ylim=c(0.0,.75),xlim=c(-10,150))+
  annotate(geom="text",x=-8,y=4.60,angle=90,label="Nlim Proline, +400uM Q")
u



```



```{r,eval=F,echo=F}
#load("../tmp/polyAupshift.RData")
#g<-ggplot(polyAdata)+theme_classic()+
#  aes(x=log10(FL1.A),col=RNAsed)+
#	stat_bin(aes(y=..density..),binwidth=0.05,position="identity",geom="line")+
#	stat_bin(aes(y=..density..),binwidth=0.05,position="identity",geom="point")+
#	facet_grid(Minutes~.,scales="free")+
#	ggtitle("polyA signal density, log10 scale, with negatives")+
#	scale_color_discrete("RNAsed before hybridization?")+
#	theme(legend.position="bottom",axis.text.y=element_blank())+
#	coord_cartesian(xlim=c(2.5,5.25))
#g
#ggsave("../supplement/figures/Figure1_S_polyAcontrols.png",g,width=7,height=7)
#
#figpolya<-ggplot(subset(polyAdata,RNAsed==FALSE&Minutes>=0))+theme_classic()+
#  aes(y=FL1.A,group=Minutes,x=Minutes/60)+
#  geom_boxplot()+
#  xlab("Hours")+
#  scale_x_continuous(breaks=seq(-4,7,1))+
#  ylab("log10 poly-A tail content \n( by probe signal area, AU )")+
#  theme(panel.grid.major=element_line(linetype="dotted")#color="grey70")
#    ,panel.grid.major.x=element_blank())+
#  scale_y_log10(breaks=c(1e4,2.5e4,5e4,1e5))+
#  coord_cartesian(ylim=c(7e3,6e4))
#figpolya
#figpolyaScaled <- figpolya+scale_x_continuous(breaks=seq(-4,7,2),limits=c(-4.5,6.5))
#
#g <- cowplot::plot_grid(
#  fig1b,fig1c,figpolyaScaled,labels=c("A","B","C")
#  ,ncol=1,align="v",axis="l")
#g
#ggsave("../supplement/figures/Figure1_S_allThree.png",g)
```

dme148

- On 151111, backed a 5ml O/N of FY4, 1ml onto 50ml NlimPro(800uM),
  @1408, put at 30C shaker.
- Started collecting at 0940 in the morning. At each time, sampled 
  8ml onto a filter and flash froze in liquid nitrogen.
  The first sample (-37min) was 10ml, different.
  (~1min procedure).
- @0942, sonicated 0.5ml of the culture and counted on a 
  hemacytometer as 11.4e6 cells per ml, which is late exponential.
- @1017, at the sample time as filtering the 0min sample, 
  added 128ul 100mM glutamine to the culture.
- By last sample @1419, cells were approximately 24.4 e6 cells per ml.
- Every sample was frozen in liquid nitrogen, then put into -80C.

For each sample, on 151117.

- @~1130, R/S sample with vortexing into 4% fresh PFA 0.75x PBS.
  Immediately discarded filter, fixed on bench RT for 3 hours.
- Added 200ul 2.5M glycine to quench, then spun 6000g RT 1min.
- Aspirated, washed and spun twice with 1ml 1x PBS.
- R/S with 898ul of Buffer B (100mM potassium phosphate buffer, 1.2M
  sorbitol, pH ~7.4), plus 100ul 200mM vanadyl ribonucleoside complex,
  0.2% beta-mercaptoethanol, and 100 units lyticase 
  (sigma, R/S in 1x PBS and stored at -20C in aliquots). 
  Incubated this at 37C for an hour. 
- Spun 3min 1200g, washed with cold Buffer B, spun and washed twice.
- R/S with 1ml 70% etOH, into 4C.

On 151118:

- In afternoon, took two 400ul of each sample. First aliquot spun
  and R/S in 1x TE with 200ug/ml RNAseA, incubated 42C for 30min.
  For the second, ommited this step.
- Spun all, aspirated, R/S in 1ml 2x SSC.
- Spun all, aspirated, R/S in 100ul of hybridization buffer with
  RNAsed as mastermixed for all samples
  (10% dextran sulfate w/v, 2x SSC final, 100ug/ml ecoli tRNA, 
  250mg/ml salmon sperm DNA) plus 100nM of an oligo, (dT)50+V 
  5'-labeled with with Alexa 488, as ordered from IDT. 
- Incubated ~14.5 hours on a 37C roller drum.

On 151119:

- Added 1ml 2x SSC and spun 10min 800g. 
- Aspirated, R/S 1ml 2x SSC and incubated 37C 15min.
- Spun 5min, aspirated, R/S 2x SSC 1ml.
- Spun 5min, aspirated, R/S in 200ul filtered 1x PBS. Sonicated as for
  the coulter counter. Kept on ice.
- Flowed on a BD Accuri, one at a time, with flicking before the run.




# Poly(A) content determination

Purpose: to measure the polyA content of cells grown in different
conditions, in order to validate some of Niki's measurements.

Chemostat cultures were established under carbon and 
nitrogen source limitation at three dilution rates. When stable, 
5ml of cultures were filtered and flash-frozen 
with liquid $N_2$. Samples were re-suspended in 1ml 4% 
paraformaldehyde (EMS) buffered in 0.75x PBS and incubated at 
RT for 2 hours before quenching with 1/5th volume 2.5M glycine. 
After washing with Buffer B (1.2M sorbitol 100mM $KHPO_4$ pH 7.5),
samples were digested with 100U lyticase (Sigma), in 1ml of 
Buffer B with 20mM VRC (NEB) and 28.6mM Beta-mercaptoethanol, 
for one hour at 37C before centrifuging at 1200g and washing 
with Buffer B.
Cells were resuspended in 70% ethanol at 4C overnight to 
permeabilize. Cells were centrifuged at 1200g, washed with 2x SSC,
and resuspended in hybridization buffer (4x SSC, 
100mg/ml dextran sulfate (Sigma), 100ug/ml E. coli tRNA (Roche), 
250ug/ml salmon sperm DNA) with 100nM (dT)$_{50}$V oligos labeled 
singly on the 5' with Alexa488 (IDT), 
and incubated overnight (~16hr) at 37C.
Cells were washed thrice with 2x SSC, resuspended in filtered 1x PBS,
and sonicated before being run on a BD Accuri flow cytometer.
Events were analyzed in R 
to gate against doublets on forward-scatter area vs. 
height, gated for size with forward-scatter area 
threshold for each growth rate, and
poly(A) content signal was determined by the signal area on the
514/20nm detector.

(see below)


```{r,warning=F,cache=T}

if (length(dir("polyeh/exp154polyA/"))) {
  directories <- c(
    exp154polyA="polyeh/exp154polyA/")
}
datz <- list()
for (expr in names(directories)) {
  dirz <- directories[expr]
  for (fz in dir(dirz)[grep("fcs",dir(dirz))]) {
    if (grepl("(water)|(h2o)|(decon)|(pbs)",fz)) {next}
    print(paste0("doing ",dirz,"",fz))
    datz[[paste0(expr, warning=F,"_",sub(".fcs","",fz))]] <- data.frame(
      exprs(read.FCS(paste0(dirz,fz),transformation=FALSE)))
  }
}
if (exists("lrgdf")) {rm(lrgdf)}
for (fz in names(datz)) {
  relevantBit <- gsub(".+?\\s(.+?)$","\\1",fz)
  if (!exists("lrgdf")) {
    lrgdf <- data.frame(datz[[as.character(fz)]],
      reshape2::colsplit(relevantBit,"_",
        names=c("sample","polydTProbe")))
  } else {
    lrgdf <- rbind(lrgdf,data.frame(datz[[as.character(fz)]],
      reshape2::colsplit(relevantBit,"_",
        names=c("sample","polydTProbe")))
    )
  }
}
dim(lrgdf)

lrgdf$nutrientLimitation <- factor(ifelse(
  grepl("[g1][123]",lrgdf$sample),
    "NitrogenLimitation","CarbonLimitation"))
lrgdf$dilutionRate[grepl("[g1][14]",lrgdf$sample)] <- "0.12"
lrgdf$dilutionRate[grepl("[g1][25]",lrgdf$sample)] <- "0.20"
lrgdf$dilutionRate[grepl("[g1][36]",lrgdf$sample)] <- "0.30"
lrgdf$dilutionRate <- factor(lrgdf$dilutionRate)
lrgdf$chemostat <- factor(ifelse(
  grepl("g[1-6]",lrgdf$sample),"Gseries","1series"))

```

Here we take out the negatives until gating is resolved.

First downsample everything to help with the gating, 
with a downsampled data.frame.

```{r,warning=F,cache=T}
negdf <- subset(lrgdf,polydTProbe=="-")
lrgdf <- subset(lrgdf,polydTProbe=="+")

for (samz in unique(lrgdf$sample)) {
  if (!exists("dsdf")) {
    dsdf <- data.frame(subset(lrgdf,sample==samz)[
      sample(1:sum(lrgdf$sample==samz),1e4),])
  } else {
    dsdf <- rbind(dsdf,data.frame(subset(lrgdf,sample==samz)[
      sample(1:sum(lrgdf$sample==samz),1e4),]))
  }
}
```

First, cut out doublets:

```{r,warning=F,cache=T}
ggplot(dsdf)+theme_bw()+
  aes(x=log(FSC.A),y=log(FSC.H))+
  facet_grid(dilutionRate~nutrientLimitation+chemostat)+
  geom_point(size=.1,alpha=.01)+
  geom_abline(slope=0.9,intercept=1.55,linetype="dotted")
dsdf <- subset(dsdf,log(FSC.H)>(log(FSC.A)*.9+1.55))
```

Then let's look at FSC and SSC. Sizes look a little odd, there's a
trend of expansion in the G456 and 1123, but not the others.
Reminescent of the split in the sytoxgreen.

```{r,warning=F,cache=T}
ggplot(dsdf)+theme_bw()+
  aes(x=log(FSC.A),y=log(SSC.A))+
  facet_grid(dilutionRate~nutrientLimitation+chemostat)+
  geom_point(size=.1,alpha=.01)+
  geom_vline(xintercept=log(c(6e4,4e4,2e4)),linetype="dotted")
dsdf <- subset(dsdf,
  (dilutionRate=="0.30"&(FSC.A)>6e4)|
  (dilutionRate=="0.20"&(FSC.A)>4e4)|
  (dilutionRate=="0.12"&(FSC.A)>2e4)
)
```

```{r,warning=F,cache=T}
subdf <- subset(lrgdf,
  log(FSC.H)>(log(FSC.A)*.9+1.55)&
  ((dilutionRate=="0.30"&(FSC.A)>6e4)|
  (dilutionRate=="0.20"&(FSC.A)>4e4)|
  (dilutionRate=="0.12"&(FSC.A)>2e4))
)
snegdf <- subset(negdf,
  log(FSC.H)>(log(FSC.A)*.9+1.55)&
  ((dilutionRate=="0.30"&(FSC.A)>6e4)|
  (dilutionRate=="0.20"&(FSC.A)>4e4)|
  (dilutionRate=="0.12"&(FSC.A)>2e4))
)
alldf <- rbind(subdf,snegdf)
```

Below, we subset out the chemostat series that didn't look
contaminated in the sytoxgreen DNA content measures, then look
at their distributions and statistics.

First, the positives and negatives (no probe).

```{r,warning=F,cache=T}
g2<-ggplot(subset(alldf,sample%in%c("g4","g5","g6","11","12","13")))+
  theme_bw()+
  aes(x=log10(FL1.A),col=polydTProbe)+
  facet_grid(dilutionRate~nutrientLimitation)+
  stat_bin(aes(y = ..density..),binwidth=0.010,geom="line",
    position="identity",cex=.5)+
  stat_bin(aes(y = ..density..),binwidth=0.010,geom="point",
    position="identity",cex=.2)+
  theme(legend.position="bottom")+
  coord_cartesian(xlim=c(2.0,5.5))+
  ylab("Density")+theme(axis.text.y=element_blank())+
  xlab("log10( poly-dT probe Alexa488 signal )")
g2
```

Okay, clear difference, below is just the positives.

Here's a boxplot.

```{r,warning=F,cache=T}
g3<-ggplot(subset(alldf,sample%in%c("g4","g5","g6","11","12","13")&
    polydTProbe=="+"))+theme_bw()+
  aes(x=as.numeric(as.character(dilutionRate)),group=dilutionRate,
    y=log10(FL1.A))+
  facet_wrap(~nutrientLimitation)+
  geom_boxplot(width=0.075,outlier.size=0.0001)+
  theme(axis.text.x=element_text(angle=90))+
  coord_cartesian(ylim=c(3.25,5.25))+
  xlab("Dilution Rate, volumes per hour")+
  ylab("log10( poly-dT probe Alexa488 signal )")
g3
```
```{r,cache=F}
ggsave("../img/polya_niki_box.png",g3,width=4,height=4)
```

But just the descriptive stats would be nice.

```{r,warning=F,cache=T}
descStatz <- plyr::ddply(alldf,
  c("polydTProbe","nutrientLimitation","dilutionRate","chemostat"),
  function(x){
    return(c(mean=mean(x[["FL1.A"]]),median=median(x[["FL1.A"]]),
      quantile(x[["FL1.A"]],probs=c(0.25,0.75)))
    )
  })
```

So here's that, plotted:

```{r,warning=F,cache=T}
g6 <- ggplot(reshape2::melt(subset(descStatz,polydTProbe%in%c("+")&
     ((chemostat=="1series"&nutrientLimitation=="NitrogenLimitation")|
     (chemostat=="Gseries"&nutrientLimitation=="CarbonLimitation"))),
    measure.vars=c("mean","median","25%","75%"),
    id.vars=c("nutrientLimitation",
      "dilutionRate","chemostat")))+
  theme_bw()+
  aes(x=as.numeric(as.character(dilutionRate)),y=log10(value),
    linetype=variable)+
  scale_linetype_manual("Summary statistic",
    values=c("dashed","solid","dotted","dotted"),
    labels=c("Mean","Median","25% quantile","75% quantile"))+
  facet_wrap(~nutrientLimitation)+
  theme(legend.position="bottom")+
  geom_point(size=0.5)+geom_line()+
  xlab("Dilution rate, volumes per hour")+
  ylab("log10( poly-dT Alexa488 probe signal )")
g6
```

Here's the actual numbers, for all measured, negatives too.

```{r,warning=F,cache=F}
descStatz
```

And for the subset of ones actually used:

```{r,warning=F,cache=F}
tmp <- subset(descStatz,polydTProbe%in%c("+")&
  ((chemostat=="1series"&nutrientLimitation=="NitrogenLimitation")|
  (chemostat=="Gseries"&nutrientLimitation=="CarbonLimitation")))
tmp[,-1]
```

And how the mean polyA content signal compares to the totals
of the mean columns in Tables S2 and S3, the line is a `lm()` call
for linear dependence.

Note that the slow growing carbon outlier is G4, 
and take a look at G4 in the sytoxgreen data. It's contaminated.
Just saying.

```{r,warning=F,cache=T,echo=F}
compdat <- data.frame(
  tmp[,c("nutrientLimitation","dilutionRate","mean")],
  sparqTotalMoleculesPerCell=c(3866.90,6334.23,8506.86,
    995.526,2987.95,9587.64))
g10 <- ggplot(compdat)+theme_bw()+
  aes(x=sparqTotalMoleculesPerCell,y=mean)+
  geom_point(aes(col=nutrientLimitation))+
  xlab("Mean mRNA per cell,\nabsolute measurement,\nby SPARQ")+
  ylab("Mean mRNA signal per cell,\nrelative measurement,\nby poly-dT hybridization")+
  stat_smooth(method="lm",se=F,col="black",linetype="dashed",
    lwd=0.1,fullrange=T)+
  coord_cartesian(xlim=c(0,1e4),ylim=c(0,5e4))+
  guides(col=guide_legend(title="Nutrient limitation",ncol=1))+
  theme(legend.position="bottom")
g10
```
```{r,cache=F}
ggsave("../img/polya_niki_summary.png",g10,
  width=4,height=4)
```
```{r,cache=T}

summary(lm(compdat$sparqTotalMoleculesPerCell~compdat$mean))
cor.test(compdat$sparqTotalMoleculesPerCell,compdat$mean)
```

Without the contaminated G4 sample?

```{r,warning=F,cache=T,echo=F}
ggplot(compdat[-1,])+theme_bw()+
  aes(x=sparqTotalMoleculesPerCell,y=mean)+
  geom_point(aes(col=nutrientLimitation))+
  xlab("Mean mRNA per cell,\nby SPARQ")+
  ylab("Mean mRNA per cell,\nby poly-dT hybridization")+
  stat_smooth(method="lm",se=F,col="black",linetype="dashed",
    lwd=0.1,fullrange=T)+
  coord_cartesian(xlim=c(0,1e4),ylim=c(0,5e4))
summary(lm(compdat$sparqTotalMoleculesPerCell[-1]~compdat$mean[-1]))
```

Oh well.

