
```{r}
library(tidyverse)
library(limma)
library(reshape2)
```

```{r,cache=T}
#datadir <- "~/lab/chunks/heatshockStephensProject/data/"
#mdatar <- read.csv(paste0(datadir,"countedDataForAnalysissne26.csv"))
mdatar <- read.csv("sne26barseqHeatshockStephen/countedDataForAnalysissne26.csv")
datar <- dcast(subset(mdatar,Tag=="UP"),Strain~SampleNum+whichRun+Q+time+replicate,value.var="Counts")
rownames(datar) <- datar[,1]
datar <- datar[,-1]
```

That read in the data. If you are not Darach and you're running this,
you'll have to get your data in somewhere else.


How do the counts look? So we're looking across each row of genes,
and we're asking how many counts do we have?

```{r}
library(tidyverse)
plot(density(apply(datar,1,function(x){sum(x)}),0.1))

g <- datar %>% as.tibble() %>% gather(Variable,Value) %>% 
  separate(Variable,into=c("Sample","Run","Shifted","Shock","Rep"))%>%
  mutate(Shock=factor(Shock,levels=c("preShift","preShock","postShock")))%>%
  arrange(Shock)%>%
  mutate(Shock=c(`preShift`="Before upshift",`preShock`="After shift\nbefore shock",`postShock`="After shock\nand outgrowth")[Shock])%>%
  mutate(Shock=factor(Shock,levels=unique(Shock)))%>%
  ggplot()+theme_bw()+aes(x=Value)+
  facet_grid(Shock~.)+
  geom_histogram(bins=50)+
  scale_x_log10(limits=c(0.5,1e3))+
  ylab("Occurrences")+
  xlab("Counts per mutant per sample")
g
ggsave("../img/sne_histogram.png",g,width=5,height=3)

g <- datar %>% as.tibble() %>% mutate(Gene=rownames(datar)) %>%
  gather(Variable,Value,-Gene) %>% 
  separate(Variable,into=c("Sample","Run","Shifted","Shock","Rep")) %>%
  mutate(Shock=factor(Shock,levels=c("preShift","preShock","postShock")))%>%
  arrange(Shock)%>%
  mutate(Shock=c(`preShift`="Before upshift",`preShock`="After shift\nbefore shock",`postShock`="After shock\nand outgrowth")[Shock])%>%
  mutate(Shock=factor(Shock,levels=unique(Shock)))%>%
  group_by(Shock,Gene)%>%summarize(Value=sum(Value),na.rm=T)%>%
  ggplot()+theme_bw()+aes(x=Value)+
  facet_grid(Shock~.)+
  geom_histogram(bins=50)+
  scale_x_log10(limits=c(0.5,1e4))+
  ylab("Occurrences")+
  xlab("Counts per mutant, across samples")
g
ggsave("../img/sne_histogram_sum.png",g,width=5,height=3)

```

Basically, I looped between this and looking at the mean vs var
plot from limma, and picked this cut off

```{r,cache=T}
plot(density(apply(datar,1,function(x){sum(x)})[apply(datar,1,function(x){sum(x)})>1],0.1))
plot(density(apply(datar,1,function(x){sum(x)})[apply(datar,1,function(x){sum(x)})>5],0.1))
plot(density(apply(datar,1,function(x){sum(x)})[apply(datar,1,function(x){sum(x)})>15],0.1))
plot(density(apply(datar,1,function(x){sum(x)})[apply(datar,1,function(x){sum(x)})>30],0.1))
plot(density(apply(datar,1,function(x){sum(x)})[apply(datar,1,function(x){sum(x)})>50],0.1))
plot(density(apply(datar,1,function(x){sum(x)})[apply(datar,1,function(x){sum(x)})>70],0.1))
plot(density(apply(datar,1,function(x){sum(x)})[apply(datar,1,function(x){sum(x)})>100],0.1))
sdatar <- list()
sdatar[["0"]] <- datar[apply(datar,1,function(x){sum(x)})>0,]
sdatar[["15"]] <- datar[apply(datar,1,function(x){sum(x)})>15,]
sdatar[["30"]] <- datar[apply(datar,1,function(x){sum(x)})>30,]
sdatar[["70"]] <- datar[apply(datar,1,function(x){sum(x)})>70,]
```

Here we build the design matrix.

```{r,cache=T}
d <- dcast(subset(mdatar,Tag=="UP"),
  SampleNum+whichRun+Q+time+replicate~.,value.var="Counts")
shifted <- relevel(factor(!grepl("preShift",d$time)),"FALSE")
shocked <- relevel(factor(grepl("postShock",d$time)),"FALSE")
whichRun <- factor(d$whichRun)
Q <- relevel(factor(d$Q),"FALSE")
design <- model.matrix(~0+whichRun+shifted+shifted:Q+shocked+shocked:Q)
```

```{r,cache=T}

voomer <- function(x,name) {
  vdatar <- voom(x,design=design,plot=T,save.plot=T)
  fit <- lmFit(vdatar,design)
  return(list(name,vdatar,fit))
}

z<-imap(sdatar,voomer) %>%tibble(RawOut=.) %>% 
  mutate(Threshold=map(RawOut,~.x[[1]]))%>%
  mutate(VoomDatar=map(RawOut,~.x[[2]]))%>%
  mutate(VoomFit=map(RawOut,~.x[[3]]))%>%
  unnest(Threshold)%>%
  dplyr::select(-RawOut)

```

```{r,cache=T}

pdat <- z%>%
  mutate(pointz_x=map(VoomDatar,~.x[["voom.xy"]][[1]]))%>%
  mutate(pointz_y=map(VoomDatar,~.x[["voom.xy"]][[2]]))%>%
  mutate(linez_x=map(VoomDatar,~.x[["voom.line"]][[1]]))%>%
  mutate(linez_y=map(VoomDatar,~.x[["voom.line"]][[2]]))%>%
  dplyr::select(Threshold,ends_with("_x"),ends_with("_y"))%>% unnest() %>%
  gather(Variable,Value,ends_with("_x"),ends_with("_y"))%>%
  separate(Variable,into=c("Type","Axis"),sep="_") %>%
  nest(Value) %>% spread(Axis,data) %>%unnest(x,y)%>%
  dplyr::rename(x=Value,y=Value1)%>%
  arrange(Threshold,x)%>%
  filter(Threshold%in%c("0","30"))

g <- ggplot()+theme_bw()+
  facet_wrap(~Threshold)+
  geom_point(data=pdat%>%filter(Type=="pointz")
    ,aes(x=x,y=y),size=0.1,alpha=0.5)+
  geom_line(data=pdat%>%filter(Type=="linez")
    ,aes(x=x,y=y,col=Threshold))+
  xlab("Abundance")+ylab("Variance")
g
ggsave("../img/sne_voomer.png",g,width=7,height=3)

```

Then pick contrasts to look at, basically that's a vector of values
that show which weights we're testing for difference associated.

```{r}
data.frame(colnames(design))
```

Which are increased in the second run?

Just gonna use the cutoff of more than 30 counts total to filter.

```{r}
fit <- z%>%filter(Threshold=="30")%>%pull(VoomFit)%>%{.[[1]]}
runDiff <- eBayes(contrasts.fit(fit,c(-1,1,0,0,0,0,0)))
runDiffTable <- topTable(runDiff,sort.by="logFC",number=10000)
runDiffTable[order(runDiffTable$logFC,decreasing=T),][1:20,]
```
Which are statistically significantly increased in the second run?
```{r}
subset(runDiffTable[order(runDiffTable$logFC,decreasing=T),],adj.P.Val<0.05)
```


Which are increased in the glutamine flasks regardless of treatment?
```{r}
flaskDiff <- eBayes(contrasts.fit(fit,c(0,0,0,0,1,0,0)))
flaskDiffTable <- topTable(flaskDiff,sort.by="logFC",number=10000)
flaskDiffTable[order(flaskDiffTable$logFC,decreasing=T),][1:20,]
```
Which are statistically significantly increased in the glutamine flasks regardless of treatment?
```{r}
subset(flaskDiffTable[order(flaskDiffTable$logFC,decreasing=T),],adj.P.Val < 0.05)
```

Which are increased in shifting? 
```{r}
shiftDiff <- eBayes(contrasts.fit(fit,c(0,0,1,0,0,0,0)))
shiftDiffTable <- topTable(shiftDiff,sort.by="logFC",number=10000)
shiftDiffTable[order(shiftDiffTable$logFC,decreasing=T),][1:20,]
```
Which are statistically significantly increased in shifting?
```{r}
subset(shiftDiffTable[order(shiftDiffTable$logFC,decreasing=T),],adj.P.Val < 0.05)
```


Which are increased in glutamine shifting? 
```{r}
qshiftDiff <- eBayes(contrasts.fit(fit,c(0,0,0,0,0,1,0)))
qshiftDiffTable <- topTable(qshiftDiff,sort.by="logFC",number=10000)
qshiftDiffTable[order(qshiftDiffTable$logFC,decreasing=T),][1:20,]
```
Which are statistically significantly increased in glutamine shifting?
```{r}
subset(qshiftDiffTable[order(qshiftDiffTable$logFC,decreasing=T),],adj.P.Val < 0.05)
```

Which are increased in shocking? 
```{r}
shockDiff <- eBayes(contrasts.fit(fit,c(0,0,0,1,0,0,0)))
shockDiffTable <- topTable(shockDiff,sort.by="logFC",number=10000)
subset(shockDiffTable[order(shockDiffTable$logFC,decreasing=T),],adj.P.Val<0.05)
dim(subset(shockDiffTable[order(shockDiffTable$logFC,decreasing=T),],adj.P.Val<0.05))
dim(subset(shockDiffTable[order(shockDiffTable$logFC,decreasing=T),],adj.P.Val<0.05&logFC>0))
```

Which are increased in shocked glutamine flasks? 
```{r}
qshockDiff <- eBayes(contrasts.fit(fit,c(0,0,0,0,0,0,1)))
qshockDiffTable <- topTable(qshockDiff,sort.by="logFC",number=10000)
subset(qshockDiffTable[order(qshockDiffTable$logFC,decreasing=T),],adj.P.Val<0.05)
dim(subset(qshockDiffTable[order(qshockDiffTable$logFC,decreasing=T),],adj.P.Val<0.05))
dim(subset(qshockDiffTable[order(qshockDiffTable$logFC,decreasing=T),],adj.P.Val<0.05&logFC>0))

load("NameIDList.RData")
zisone <- qshockDiffTable%>%as.tibble()%>%
  mutate(Systematic=rownames(qshockDiffTable))%>%
  mutate(Common=renameSystematicToCommon[Systematic])%>%
  dplyr::select(Systematic,Common,logFC,t,adj.P.Val)%>%
  arrange(-t)
```

How robust to thresholds?

```{r}
fit <- z%>%filter(Threshold=="0")%>%pull(VoomFit)%>%{.[[1]]}
qshockDiff00 <- eBayes(contrasts.fit(fit,c(0,0,0,0,0,0,1)))
qshockDiff00Table <- topTable(qshockDiff00,sort.by="logFC",number=10000)
subset(qshockDiff00Table[order(qshockDiff00Table$logFC,decreasing=T),],adj.P.Val<0.05)
```

```{r}
fit <- z%>%filter(Threshold=="15")%>%pull(VoomFit)%>%{.[[1]]}
qshockDiff15 <- eBayes(contrasts.fit(fit,c(0,0,0,0,0,0,1)))
qshockDiff15Table <- topTable(qshockDiff15,sort.by="logFC",number=10000)
subset(qshockDiff15Table[order(qshockDiff15Table$logFC,decreasing=T),],adj.P.Val<0.05)
```

```{r}
fit <- z%>%filter(Threshold=="30")%>%pull(VoomFit)%>%{.[[1]]}
qshockDiff30 <- eBayes(contrasts.fit(fit,c(0,0,0,0,0,0,1)))
qshockDiff30Table <- topTable(qshockDiff30,sort.by="logFC",number=10000)
subset(qshockDiff30Table[order(qshockDiff30Table$logFC,decreasing=T),],adj.P.Val<0.05)
```

```{r}
fit <- z%>%filter(Threshold=="70")%>%pull(VoomFit)%>%{.[[1]]}
qshockDiff70 <- eBayes(contrasts.fit(fit,c(0,0,0,0,0,0,1)))
qshockDiff70Table <- topTable(qshockDiff70,sort.by="logFC",number=10000)
subset(qshockDiff70Table[order(qshockDiff70Table$logFC,decreasing=T),],adj.P.Val<0.05)
```


```{r,clusterProfilerSetup,cache=T}
library(clusterProfiler)
library(org.Sc.sgd.db)

QueryHighEntrezIDs <- zisone%>%filter(logFC>0&adj.P.Val<0.05)%>%
  mutate(SGD=renameSystematicToSGDID[Systematic])%>%pull(SGD)%>%
  bitr(.,fromType="SGD",toType="ENTREZID"
    ,OrgDb=org.Sc.sgd.db,drop=T) %>%
  pull(ENTREZID)

QueryLowEntrezIDs <- zisone%>%filter(logFC>0&adj.P.Val<0.05)%>%
  mutate(SGD=renameSystematicToSGDID[Systematic])%>%pull(SGD)%>%
  bitr(.,fromType="SGD",toType="ENTREZID"
    ,OrgDb=org.Sc.sgd.db,drop=T) %>%
  pull(ENTREZID)

BackgroundEntrezIDs <- zisone%>%filter(adj.P.Val>0.05)%>%
  mutate(SGD=renameSystematicToSGDID[Systematic])%>%pull(SGD)%>%
  bitr(.,fromType="SGD",toType="ENTREZID"
    ,OrgDb=org.Sc.sgd.db,drop=T) %>%
  pull(ENTREZID)

read_tsv("sgd_go_full_171013.txt",col_names=F
  ,comment="!") -> SGDGOFull

read_tsv("sgd_go_terms_171013.txt",col_names=F
  ,comment="!") %>%
  mutate(GOID=str_c("GO:"
    ,str_pad(string=X1,width=7,side="left",pad="0"))
    ) -> SGDGOTerms

sysToSGD <- function(x) renameSystematicToSGDID[x]
sgdToCommon <- function(x) renameSystematicToCommon[
  names(renameSystematicToSGDID[which(renameSystematicToSGDID==x)])]
zisone <- zisone %>% 
  mutate(SGD=unlist(map(Systematic,sysToSGD)))

nPermutations <- 1e7
```

```{r,cache=T}

CategoricalGO_High_MF <- enrichGO(gene=QueryHighEntrezIDs
  ,universe=BackgroundEntrezIDs
  ,OrgDb=org.Sc.sgd.db
  ,ont="MF"#MF BP CC all
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.05
  ,qvalueCutoff=0.05
  ,minGSSize=2,maxGSSize=500)
CategoricalGO_High_MF@result

CategoricalGO_High_BP <- enrichGO(gene=QueryHighEntrezIDs
  ,universe=BackgroundEntrezIDs
  ,OrgDb=org.Sc.sgd.db
  ,ont="BP"#MF BP CC all
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.05
  ,qvalueCutoff=0.05
  ,minGSSize=2,maxGSSize=500)
CategoricalGO_High_BP@result

CategoricalGO_High_CC <- enrichGO(gene=QueryHighEntrezIDs
  ,universe=BackgroundEntrezIDs
  ,OrgDb=org.Sc.sgd.db
  ,ont="CC"#MF BP CC all
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.05
  ,qvalueCutoff=0.05
  ,minGSSize=2,maxGSSize=500)
CategoricalGO_High_CC@result

CategoricalKEGG_High <- zisone%>%filter(logFC>0&adj.P.Val<0.05)%>%
  pull(Systematic)%>%
  enrichKEGG(gene=.
    ,organism="sce"
    ,pvalueCutoff=0.05
    ,pAdjustMethod="fdr"
    )
CategoricalKEGG_High@result

CategoricalGO_Low_MF <- enrichGO(gene=QueryLowEntrezIDs
  ,universe=BackgroundEntrezIDs
  ,OrgDb=org.Sc.sgd.db
  ,ont="MF"#MF BP CC all
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.05
  ,qvalueCutoff=0.05
  ,minGSSize=2,maxGSSize=500)
CategoricalGO_Low_MF@result

CategoricalGO_Low_BP <- enrichGO(gene=QueryLowEntrezIDs
  ,universe=BackgroundEntrezIDs
  ,OrgDb=org.Sc.sgd.db
  ,ont="BP"#MF BP CC all
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.05
  ,qvalueCutoff=0.05
  ,minGSSize=2,maxGSSize=500)
CategoricalGO_Low_BP@result

CategoricalGO_Low_CC <- enrichGO(gene=QueryLowEntrezIDs
  ,universe=BackgroundEntrezIDs
  ,OrgDb=org.Sc.sgd.db
  ,ont="CC"#MF BP CC all
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.05
  ,qvalueCutoff=0.05
  ,minGSSize=2,maxGSSize=500)
CategoricalGO_Low_CC@result

CategoricalKEGG_Low <- zisone%>%filter(logFC<0&adj.P.Val<0.05)%>%
  pull(Systematic)%>%
  enrichKEGG(gene=.
    ,organism="sce"
    ,pvalueCutoff=0.05
    ,pAdjustMethod="fdr"
    )
CategoricalKEGG_Low@result

```



```{r,moreSurival,dependson="clusterProfilerSetup",cache=T}
The_Input <- zisone %>% 
  mutate(SortMetric=t)%>%
  arrange(-SortMetric) %>% filter(!is.na(SortMetric)) %>%
  {setNames(.$SortMetric,nm=.$SGD)} 
The_Results <- list()

system.time(
The_Results[["BP"]] <- GSEA(
  gene=The_Input
  ,TERM2GENE=SGDGOFull%>%filter(X9=="P")%>%dplyr::select(X5,X2)
  ,TERM2NAME=SGDGOTerms%>%dplyr::select(GOID,X2)
  ,minGSSize=2,maxGSSize=50
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.2,seed=180201
  ,nPerm=nPermutations
  )
)

system.time(
The_Results[["CC"]] <- GSEA(
  gene=The_Input
  ,TERM2GENE=SGDGOFull%>%filter(X9=="C")%>%dplyr::select(X5,X2)
  ,TERM2NAME=SGDGOTerms%>%dplyr::select(GOID,X2)
  ,minGSSize=2,maxGSSize=50
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.2,seed=180201
  ,nPerm=nPermutations
  )
)

system.time(
The_Results[["MF"]] <- GSEA(
  gene=The_Input
  ,TERM2GENE=SGDGOFull%>%filter(X9=="F")%>%dplyr::select(X5,X2)
  ,TERM2NAME=SGDGOTerms%>%dplyr::select(GOID,X2)
  ,minGSSize=2,maxGSSize=50
  ,pAdjustMethod="fdr"
  ,pvalueCutoff=0.2,seed=180201
  ,nPerm=nPermutations
  )
)

```

```{r,dependson="moreSurvival",cache=T}
map(The_Results,~.x@result%>%arrange(-enrichmentScore))
```
