\documentclass[12pt,letterpaper]{memoir}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}

\usepackage[utf8]{inputenc}
\usepackage{DejaVuSansCondensed}
\renewcommand\familydefault{\sfdefault} 
\usepackage[T1]{fontenc}

\usepackage{wrapfig}
\usepackage{framed}

\usepackage{pdfpages}

\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
  \expandafter\origfigure\expandafter[H]
  } { \endorigfigure }

\usepackage{amsmath}

\usepackage{tikz}
\usepackage{pgfmath}
%\usepackage{pgffor}
\usetikzlibrary{arrows,positioning,shapes,patterns,decorations} 
\tikzstyle{every picture}+=[remember picture]

\usepackage[backend=bibtex8%biber%bibtex8
  ,backref=true,hyperref=true
  ,refsection=none %  ,refsection=chapter
  ,style=authoryear-comp]{biblatex}
\addbibresource{references_dissertation_darachmiller.bib}

\usepackage{hyperref}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black, urlcolor=black }

\makeatletter 
\def\makemytitle{
  \newpage
\begin{center}
\vfill
  \@title\\
\vfill
  by\\
\vfill
  \@author\\
\vfill
  A dissertation submitted in partial fulfillment \\
  of the requirements for the degree of \\
  Doctor of Philosophy \\
  Department of Biology \\
  New York University \\
\vfill
  \monthz, \yearz\\
  \end{center}
\vfill
\begin{flushright}
  $\rule{5cm}{0.01mm}$\\
  David Gresham
\end{flushright}
  \pagebreak
\begin{center}
  \textcopyright \hspace{0.2em} \@author \\
  All rights reserved, \yearz \\
\end{center}
  \pagebreak
}
\makeatother

\title{The extent and factors of rapid repression via mRNA
destabilization upon the glutamine upshift in budding yeast.}
\author{Darach Miller}
\def\monthz{May}
\def\yearz{2018}

%
%
%
%
%
%

\includeonly{%dissertation_darachmiller_1,dissertation_darachmiller_2,
dissertation_darachmiller_3}
\hfuzz=20pt
\vfuzz=20pt
\hbadness=2000
\vbadness=\maxdimen

\setsecnumdepth{subsection}
\settocdepth{subsection}

\begin{document}
\DoubleSpacing
\frontmatter
\pagestyle{empty}
%\renewcommand{\thefigure}{\arabic{figure}}

\makemytitle

%\begin{quote}
%\SingleSpace
%"Science is a match that man has just got alight. He thought he 
%was in a room --- in moments of devotion, a temple --- and 
%that his light would be reflected from and display walls inscribed 
%with wonderful secrets and pillars carved with philosophical 
%systems wrought into harmony. 
%\vfill
%It is a curious sensation, now that the preliminary splutter is 
%over and the flame burns up clear, to see his hands lit and just 
%a glimpse of himself and the patch he stands on visible, and 
%around him, in place of all that human comfort and beauty he 
%anticipated, darkness still."
%
%\hfill - H.G. Wells, 1891
%\end{quote}

\newpage

\section*{Acknowledgements}
\addcontentsline{toc}{section}{\numberline{}Acknowledgements}

so that happened

% the collaborative environment at NYU CGSB, 
%
% Viji Subramanian, Andreas Hochwagen and his group for microscope
% access and tolerating me playing with the Deltavision.
%
% Ken Birnbaum and his group for helpful conversations and
% access to the lucky FISH incubator.
%
% Evelina Tutucci for demonstrating the fiddly FISH method.
% Megan McClean for openly sharing a FISH protocol.
%
% Carole Hom, Ian Korf, Patrice Kohel, Nitin Baliga, Aaron Brooks,
% but most importantly Marc Facciotti for giving me a chance when
% I had given mine up.
%
% Sarah Nguyen. For keeping me together, like 
%
% Michi, Henri(ci), and Tai for making me get up and come home on time.
%
%Harry Miller, Bruce Sterling, William Gibson for inspiring
%me to wander this way.
%Sue Miller for devising a clever genetic screen for yeast
%strains that fail to re-initiate growth programs.
%My sister Kendra Miller for egging me on.
%
%Staff at eBioscience/Affymetrix/ThermoFisher for support with
%the Quantigene/FlowRNA probe set and patience through the years.
%
%The Cold Spring Harbor Yeast Course, especially the instructors
%Maitreya Dunham, Marc Gartenberg, and Grant Brown for introducing
%me to the community and (re)-lighting a fire.
%
%Past students including Daniel, Alex, and Stephen.
%
%My committee for the most fun three hours I had each year.
%
%Past and present members of the Gresham and Vogel labs for 
%discussions and support.
%Especially Niki, Steff, Nathan, and Benjy; Naomi, Farah, Siyu,
%Pieter.
%

\newpage
\pagestyle{plain}

\section*{Abstract}
\addcontentsline{toc}{section}{\numberline{}Abstract}

Cellular responses to changing environments frequently involve rapid
reprogramming of the transcriptome. Regulated
changes in mRNA degradation rates can accelerate reprogramming by
clearing or stabilizing extant transcripts. 
During a nitrogen upshift, the budding yeast \textit{Saccharomyces
cerevisiae} reprograms its physiology in order to enter into  
a state of rapid growth. 
I sought to understand the role of transcriptome reprogramming
in effecting this transition.
During the first minutes of a nitrogen upshift,
preceding the changes in population growth rate,
the transcriptome quickly into a state transiently distinct from
either rapid or slow growth.
During this, the five fastest decreasing mRNA are all NCR-regulated 
transporters, suggesting that this reprogramming is specifically
targeted to set the transcriptional program for rapid growth.
Many transcripts decrease significantly faster than expected from
their stability measured during steady-state growth during
rich media conditions, so I set out to determine the stability of
the yeast transcriptome preceding and during a nitrogen upshift.
I used 4-thiouracil labeling to track mRNA stability, and found 
that 78 mRNAs are subject to destabilization. 
These transcripts include
Nitrogen Catabolite Repression (NCR) and carbon metabolism mRNAs,
suggesting that mRNA destabilization is a mechanism for targeted
reprogramming. 
To explore the molecular basis of destabilization and possible
mechanisms of specifying or effecting the destabilization, I
developed a novel method combining mRNA FISH, fluorescence-activated
cell sorting, and DNA barcode sequencing to screen the pooled deletion
collection library for \textit{trans} factors that mediate rapid GAP1 mRNA
repression. This required that I completely redesign the methods for 
low-input multiplexed barcode sequencing, as well as adapt 
branched-DNA single-molecule mRNA FISH protocols to work in budding
yeast. Modeling of the data identifies known factors of mRNA
degradation, namely all three non-essential components of the
Lsm1-7p/Pat1p complex, as being important for proper \textit{GAP1}
mRNA dynamics. However, the modulators \textit{EDC3} and \textit{SCD6}
have more complex phenotypes. Re-analyzing previously collected data,
I identified that a \textit{scd6}$\Delta$, \textit{tif4632}$\Delta$
(eIF4G2 knock-out), and a \textit{GAP1} 5' UTR delete strain all
share a similar phenotype of lower \textit{GAP1} expression preceding
the upshift, and reduced decay upon the upshift. This suggests that 
translation initiation dynamics may play a role in mRNA levels and 
their stability, and suggests that the destabilization phenomenon 
may in fact be the release of a stabilization effect.

\newpage

\tableofcontents*

\mainmatter

\include{dissertation_darachmiller_1}
\include{dissertation_darachmiller_2}
\include{dissertation_darachmiller_3}
\include{dissertation_darachmiller_4}
\include{dissertation_darachmiller_5}

\printbibliography[heading=bibintoc,title={References}]

\end{document}

\section{Acknowledgments}
We would like to acknowledge the funding source of 
NIH grant 5R01GM107466.
We would also like to thank
Andreas Hochwagen and Viji Subramanian for microscope access, 
Ken Birnbaum for helpful conversations and equipment usage, 
Michi Pedraza,
Andres Mansisidor, and Matt Paul for helpful reading and comments,
Evelina Tutucci for demonstrating FISH,
staff at 
eBioscience/Affymetrix/ThermoFisher for support with
the Quantigene/FlowRNA probe set, 
the Cold Spring Harbor Yeast Course,
the NYU Genomics Core facility for sequencing, flow cytometry,
and support,
and past and present members of the Gresham and Vogel labs for 
discussions and support.

